Future Evolution of CSMA Protocols for the IEEE 802.11 Standard

### 2013-01-25 paper.tex
fixed spaces
Boris and Azade dropbox version

%% bare_conf.tex
%% V1.3
%% 2007/01/11
%% by Michael Shell
%% See:
%% http://www.michaelshell.org/
%% for current contact information.
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.7 or later) with an IEEE conference paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
%% and
%% http://www.ieee.org/

%%*************************************************************************
%% Legal Notice:
%% This code is offered as-is without any warranty either expressed or
%% implied; without even the implied warranty of MERCHANTABILITY or
%% FITNESS FOR A PARTICULAR PURPOSE! 
%% User assumes all risk.
%% In no event shall IEEE or any contributor to this code be liable for
%% any damages or losses, including, but not limited to, incidental,
%% consequential, or any other damages, resulting from the use or misuse
%% of any information contained here.
%%
%% All comments are the opinions of their respective authors and are not
%% necessarily endorsed by the IEEE.
%%
%% This work is distributed under the LaTeX Project Public License (LPPL)
%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
%% distributed and modified. A copy of the LPPL, version 1.3, is included
%% in the base LaTeX documentation of all distributions of LaTeX released
%% 2003/12/01 or later.
%% Retain all contribution notices and credits.
%% ** Modified files should be clearly indicated as such, including  **
%% ** renaming them and changing author support contact information. **
%%
%% File list of work: IEEEtran.cls, IEEEtran_HOWTO.pdf, bare_adv.tex,
%%                    bare_conf.tex, bare_jrnl.tex, bare_jrnl_compsoc.tex
%%*************************************************************************

% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
% *** with production work. IEEE's font choices can trigger bugs that do  ***
% *** not appear when using other class files.                            ***
% The testflow support page is at:
% http://www.michaelshell.org/tex/testflow/

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print - the typesetting of the document will not typically be
% affected with changes in paper size (but the bottom and side margins will).
% Use the testflow package mentioned above to verify correct handling of
% both paper sizes by the user's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
\documentclass[conference]{IEEEtran}
% Add the compsoc option for Computer Society conferences.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[conference]{../sty/IEEEtran}

% Some very useful LaTeX packages include:
% (uncomment the ones you want to load)

% *** MISC UTILITY PACKAGES ***
%
%\usepackage{ifpdf}
% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
% compilation based on whether the output is pdf or dvi.
% usage:
% \ifpdf
%   % pdf code
% \else
%   % dvi code
% \fi
% The latest version of ifpdf.sty can be obtained from:
% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
% Also, note that IEEEtran.cls V1.7 and later provides a builtin
% \ifCLASSINFOpdf conditional that works the same way.
% When switching from latex to pdflatex and vice-versa, the compiler may
% have to be run twice to clear warning/error messages.

% *** CITATION PACKAGES ***
%
\usepackage{cite}
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
% not currently provide for hyperlinked citations.
% The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
% The documentation is contained in the cite.sty file itself.

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  % \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at: 
% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

% *** MATH PACKAGES ***
%
%\usepackage[cmex10]{amsmath}
% A popular package from the American Mathematical Society that provides
% many useful and powerful commands for dealing with mathematics. If using
% it, be sure to load this package with the cmex10 option to ensure that
% only type 1 fonts will utilized at all point sizes. Without this option,
% it is possible that some math symbols, particularly those within
% footnotes, will be rendered in bitmap form which will result in a
% document that can not be IEEE Xplore compliant!
%
% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
% thus preventing page breaks from occurring within multiline equations. Use:
%\interdisplaylinepenalty=2500
% after loading amsmath to restore such page breaks as IEEEtran.cls normally
% does. amsmath.sty is already installed on most LaTeX systems. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/

% *** SPECIALIZED LIST PACKAGES ***
%
%\usepackage{algorithmic}
% algorithmic.sty was written by Peter Williams and Rogerio Brito.
% This package provides an algorithmic environment fo describing algorithms.
% You can use the algorithmic environment in-text or within a figure
% environment to provide for a floating algorithm. Do NOT use the algorithm
% floating environment provided by algorithm.sty (by the same authors) or
% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
% algorithm float types and packages that provide these will not provide
% correct IEEE style captions. The latest version and documentation of
% algorithmic.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
% There is also a support site at:
% http://algorithms.berlios.de/index.html
% Also of interest may be the (relatively newer and more customizable)
% algorithmicx.sty package by Szasz Janos:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/

% *** ALIGNMENT PACKAGES ***
%
%\usepackage{array}
% Frank Mittelbach's and David Carlisle's array.sty patches and improves
% the standard LaTeX2e array and tabular environments to provide better
% appearance and additional user controls. As the default LaTeX2e table
% generation code is lacking to the point of almost being broken with
% respect to the quality of the end results, all users are strongly
% advised to use an enhanced (at the very least that provided by array.sty)
% set of table tools. array.sty is already installed on most systems. The
% latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/tools/

%\usepackage{mdwmath}
%\usepackage{mdwtab}
% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
% especially mdwmath.sty and mdwtab.sty which are used to format equations
% and tables, respectively. The MDWtools set is already installed on most
% LaTeX systems. The lastest version and documentation is available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools/

% IEEEtran contains the IEEEeqnarray family of commands that can be used to
% generate multiline equations as well as matrices, tables, etc., of high
% quality.

%\usepackage{eqparbox}
% Also of notable interest is Scott Pakin's eqparbox package for creating
% (automatically sized) equal width boxes - aka "natural width parboxes".
% Available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/eqparbox/

% *** SUBFIGURE PACKAGES ***
%\usepackage[tight,footnotesize]{subfigure}
% subfigure.sty was written by Steven Douglas Cochran. This package makes it
% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
% work, it is a good idea to load it with the tight package option to reduce
% the amount of white space around the subfigures. subfigure.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at:
% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
% subfigure.sty has been superceeded by subfig.sty.

%\usepackage[caption=false]{caption}
%\usepackage[font=footnotesize]{subfig}
% subfig.sty, also written by Steven Douglas Cochran, is the modern
% replacement for subfigure.sty. However, subfig.sty requires and
% automatically loads Axel Sommerfeldt's caption.sty which will override
% IEEEtran.cls handling of captions and this will result in nonIEEE style
% figure/table captions. To prevent this problem, be sure and preload
% caption.sty with its "caption=false" package option. This is will preserve
% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later 
% (recommended due to many improvements over 1.2) of subfig.sty supports
% the caption=false option directly:
%\usepackage[caption=false,font=footnotesize]{subfig}
%
% The latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
% The latest version and documentation of caption.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/

% *** FLOAT PACKAGES ***
%
%\usepackage{fixltx2e}
% fixltx2e, the successor to the earlier fix2col.sty, was written by
% Frank Mittelbach and David Carlisle. This package corrects a few problems
% in the LaTeX2e kernel, the most notable of which is that in current
% LaTeX2e releases, the ordering of single and double column floats is not
% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
% single column figure to be placed prior to an earlier double column
% figure. The latest version and documentation can be found at:
% http://www.ctan.org/tex-archive/macros/latex/base/

%\usepackage{stfloats}
% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
% the ability to do double column floats at the bottom of the page as well
% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
% LaTeX2e). It also provides a command:
%\fnbelowfloat
% to enable the placement of footnotes below bottom floats (the standard
% LaTeX2e kernel puts them above bottom floats). This is an invasive package
% which rewrites many portions of the LaTeX2e float routines. It may not work
% with other packages that modify the LaTeX2e float routines. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
% Documentation is contained in the stfloats.sty comments as well as in the
% presfull.pdf file. Do not use the stfloats baselinefloat ability as IEEE
% does not allow \baselineskip to stretch. Authors submitting work to the
% IEEE should note that IEEE rarely uses double column equations and
% that authors should try to avoid such use. Do not be tempted to use the
% cuted.sty or midfloat.sty packages (also by Sigitas Tolusis) as IEEE does
% not format its papers in such ways.

% *** PDF, URL AND HYPERLINK PACKAGES ***
%
%\usepackage{url}
% url.sty was written by Donald Arseneau. It provides better support for
% handling and breaking URLs. url.sty is already installed on most LaTeX
% systems. The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
% Read the url.sty source comments for usage information. Basically,
% \url{my_url_here}.

%\usepackage[linesnumbered,noend]{algorithm2e}
\usepackage[linesnumbered,vlined]{algorithm2e}

% *** Do not adjust lengths that control margins, column widths, etc. ***
% *** Do not use packages that alter fonts (such as pslatex).         ***
% There should be no need to do such things with IEEEtran.cls V1.6 and later.
% (Unless specifically asked to do so by the journal or conference you plan
% to submit to, of course. )

@@

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Evolution of CSMA Protocols for the IEEE 802.11 Standard}

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Luis Sanabria-Russo, Azadeh Faridi, Boris Bellalta, Jaume Barcelo, Miquel Oliver}
\IEEEauthorblockA{
Universitat Pompeu Fabra\\
Roc Boronat 183, 08018 Barcelona\\
Email: \{name.surname@upf.edu\}}
}
%\and
%\IEEEauthorblockN{Homer Simpson}
%\IEEEauthorblockA{Twentieth Century Fox\\
%Springfield, USA\\
%Email: homer@thesimpsons.com}
%\and
%\IEEEauthorblockN{James Kirk\\ and Montgomery Scott}
%\IEEEauthorblockA{Starfleet Academy\\
%San Francisco, California 96678-2391\\
%Telephone: (800) 555--1212\\
%Fax: (888) 555--1212}}

% conference papers do not typically use \thanks and this command
% is locked out in conference mode. If really needed, such as for
% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
% after \documentclass

% for over three affiliations, or if they all won't fit within the width
% of the page, use this alternative format:
% 
%\author{\IEEEauthorblockN{Michael Shell\IEEEauthorrefmark{1},
%Homer Simpson\IEEEauthorrefmark{2},
%James Kirk\IEEEauthorrefmark{3}, 
%Montgomery Scott\IEEEauthorrefmark{3} and
%Eldon Tyrell\IEEEauthorrefmark{4}}
%\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Electrical and Computer Engineering\\
%Georgia Institute of Technology,
%Atlanta, Georgia 30332--0250\\ Email: see http://www.michaelshell.org/contact.html}
%\IEEEauthorblockA{\IEEEauthorrefmark{2}Twentieth Century Fox, Springfield, USA\\
%Email: homer@thesimpsons.com}
%\IEEEauthorblockA{\IEEEauthorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
%Telephone: (800) 555--1212, Fax: (888) 555--1212}
%\IEEEauthorblockA{\IEEEauthorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}

@@

\begin{abstract}
%\boldmath
In this paper we present the requirements of candidate protocols to replace the pervasive CSMA/CA medium access control.
We discuss the possibility of further preventing collisions and provide an overview of the related work.
We specify a protocol that is a candidate for replacing CSMA/CA in pseudocode and use simulation to assess performance metrics such as throughput, fairness and collision probability.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the conference you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals/conferences frown on
% math in the abstract anyway.

% no keywords

@@

\section{Introduction}
% no \IEEEPARstart
This demo file is intended to serve as a ``starter file''
for IEEE conference papers produced under \LaTeX\ using
IEEEtran.cls version 1.7 and later.
% You must have at least 2 lines in the paragraph with the drop letter
% (should never be an issue)
I wish you the best of success.

A candidate to replace CSMA/CA should 
\begin{itemize}
  \item Provide performance advantages, either in the form of throughput or short term fairness.
  \item Be backward compatible with current implementation.
  \item Support a large number of simultaneous contenders.
  \item Be simple a simple evolution implementation to ease the transition and reduce time to market (Optional but desirable).
\end{itemize}
\section{Related Work}
Since the popularization of IEEE 802.11, several papers have proposed modifications to the contention protocol that is used for sharing the medium.
They can be categorized in three groups regarding the approach they use.
The first one prevents that the contention window is reset to its minimum value after successful transmissions.
Examples of this first group include \cite{bharghavan1994map,wang2004ncr}.
This solution improves throughput in saturation conditions at the price of lowering short term fairness.

The second groups involves the accurate estimation of the number of contenders to adjust the contention parameters.
Two examples of this group are \cite{cali2000dti,lopez-toledo2006aoi}.
This approach offers some throughput and fairness gains at the expense of increased implementation complexity.
As the number of contenders is estimated relying on the number of collisions, the presence of channel errors further complicates the estimation.
Furthermore, there is a fundamental trade-off between the accuracy and the reaction time of the estimation.

The aforementioned solutions are not able to fairly share the medium with legacy devices.
In fact, these proposals are, generally speaking, less aggressive than the currently implemented protocol.
Consequently, in a hypothetical mixed network in which the new and old protocols coexist, the new stations would receive a smaller share of the available bandwidth in a scenario.

A more important limitation of the solutions exposed so far is that the throughput is bounded by that of CSMA/CA with optimal configuration \cite{cali2000dti,bianchi2000pai}.
In the present paper we focus on a third group of solutions that delivers throughput above the maximum attainable by CSMA/CA.

This third group of solutions uses a deterministic backoff after successful transmissions to further reduce the chances of collisions.
Under certain conditions, collision-free operation is reached.
It was introduced in \cite{barcelo2008lba} and a more detailed analysis that includes both saturated and non-saturated conditions is presented in \cite{bellalta2009vtc}.
A more in-depth study is carried out in \cite{he2009srb}, including realistic elements such as clear channel assessment errors.
Different aspects of fairness are addressed in \cite{he2009srb,barcelo2010fcc,fang2011dlm}.
The performance in realistic channels taking into account the Auto Rate Fallback mechanism is evaluated \cite{martorell2012pec,martorell2012tfl}.

@@

In the present paper, we study a completely distributed solution to accommodate a large number of contenders in a fair collision-free fashion.

@@

\begin{algorithm}
\While{the device is on}
{
  %\tcc{Initialize retransmission attempts.}
  $r \leftarrow 0$ \;
  %\tcc{Initialize backoff stage.}
  $s \leftarrow 0$ \;
  %\tcc{Initialize backoff counter.}
  $b \leftarrow \mathcal{U}[0,2^s\rm{CW}_{min}-1]$\;
  \While{there is a packet to transmit}{
    %\tcc{Initialize $a$.}
    \Repeat{($r = R$) or (success)}{
      %\tcc{First, backoff.}
      \While{$b>0$}{
        wait 1 slot \;
        $b \leftarrow b-1$ \;
      }
      Attempt transmission of 1 packet \;
      \If{collision}{
        %\tcc{Random backoff.}
        $r \leftarrow r+1$ \;
        $s \leftarrow \min (s+1,S)$ \;
        $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min} -1]$\;
      }
    }
    \eIf{success}{
      %\tcc{Random backoff.}
      $r \leftarrow 0$ \;
      $s \leftarrow 0$ \;
      $b \leftarrow \mathcal{U}[0,2^{s}\rm{CW}_{min}-1]$\;
    }
    {
      Discard packet\;
      $r \leftarrow 0$ \;
      $s \leftarrow 0$ \;
      $b \leftarrow \mathcal{U}[0,2^s \rm{CW}_{min}-1]$\;
    }
  }
  Wait until there is a packet to transmit \;
}
\caption{CSMA/CA}
\label{alg:CSMA_CA}
\end{algorithm}

\begin{algorithm}
\While{the device is on}
{
  %\tcc{Initialize retransmission attempts.}
  $r \leftarrow 0$ \;
  %\tcc{Initialize backoff stage.}
  $s \leftarrow 0$ \;
  %\tcc{Initialize backoff counter.}
  $b \leftarrow \mathcal{U}[0,2^s\rm{CW}_{min}-1]$\;
  \While{there is a packet to transmit}{
    %\tcc{Initialize $a$.}
    \Repeat{($r = R$) or (success)}{
      %\tcc{First, backoff.}
      \While{$b>0$}{
        wait 1 slot \;
        $b \leftarrow b-1$ \;
      }
      Attempt transmission of 1 packet \;
      \If{collision}
      {
        %\tcc{Random backoff.}
        $r \leftarrow r+1$ \;
        $s \leftarrow \min (s+1,S)$ \;
        $b \leftarrow \mathcal{U}[0, 2^s  \rm{CW}_{min} -1]$\;
      }
    }
    \eIf{success}{
      %\tcc{Random backoff.}
      $r \leftarrow 0$ \;
      $s \leftarrow 0$ \;
      $b \leftarrow (2^{s}\rm{CW}_{min})/2$\;
    }
    {
      Discard packet\;
      $r \leftarrow 0$ \;
      $s \leftarrow 0$ \;
      $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min}-1]$\;
    }
  }
  Wait until there is a packet to transmit \;
}
\caption{CSMA/ECA}
\label{alg:CSMA_ECA}
\end{algorithm}

\begin{algorithm}
\While{the device is on}
{
  %\tcc{Initialize retransmission attempts.}
  $r \leftarrow 0$ \;
  %\tcc{Initialize backoff stage.}
  $s \leftarrow 0$ \;
  %\tcc{Initialize backoff counter.}
  $b \leftarrow \mathcal{U}[0,2^s\rm{CW}_{min}-1]$\;
  \While{there is a packet to transmit}{
    %\tcc{Initialize $a$.}
    \Repeat{($r = R$) or (success)}{
      %\tcc{First, backoff.}
      \While{$b>0$}{
        wait 1 slot \;
        $b \leftarrow b-1$ \;
      }
      Attempt transmission of 1 packet \;
      \If{collision}{
        %\tcc{Random backoff.}
        $r \leftarrow r+1$ \;
        $s \leftarrow \min (s+1,S)$ \;
        $b \leftarrow \mathcal{U}[0, 2^s  \rm{CW}_{min} -1]$\;
      }
    }
    \eIf{success}{
      %\tcc{Random backoff.}
      $r \leftarrow 0$ \;
      %$s \leftarrow 0$ \;
      $b \leftarrow (2^{s}\rm{CW}_{min})/2$\;
    }
    {
      Discard packet\;
      $r \leftarrow 0$ \;
      %$s \leftarrow 0$ \;
      $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min}-1]$\;
    }
  }
  Wait until there is a packet to transmit \;
}
\caption{CSMA/ECA with hysteresis}
\label{alg:CSMA_ECA_hist}
\end{algorithm}

>>>>

\documentclass[conference]{IEEEtran}
\usepackage{cite}
% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  % \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
\usepackage[linesnumbered,vlined]{algorithm2e}
\usepackage[ansinew]{inputenc}
\usepackage{epsfig}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{subfigure}
\usepackage[usenames]{color}
\usepackage{soul}
\usepackage{comment}
\usepackage{multirow}
\usepackage[cmex10]{amsmath}
\usepackage{bm}
\usepackage{setspace}

@@

\newcommand{\Az}[1]{{\bf{#1}}}

\begin{document}
\title{Future Evolution of CSMA Protocols for the IEEE 802.11 Standard}

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Luis Sanabria-Russo, Azadeh Faridi, Boris Bellalta, Jaume Barcelo, Miquel Oliver}
\IEEEauthorblockA{
Universitat Pompeu Fabra\\
Roc Boronat 183, 08018 Barcelona\\
Email: \{name.surname@upf.edu\}}
}

@@

\begin{abstract}
%\boldmath
In this paper we present the requirements of candidate protocols to replace the prevalent CSMA/CA medium access control.
We discuss the possibility of further preventing collisions and provide an overview of the related work.
We specify a protocol that is a candidate for replacing CSMA/CA in pseudocode and use simulation to assess performance metrics such as throughput, fairness and collision probability.
\end{abstract}

@@

\section{Introduction}
% no \IEEEPARstart
% You must have at least 2 lines in the paragraph with the drop letter
% (should never be an issue)

Since the popularization of the IEEE 802.11 standard, several works \cite{bharghavan1994map,wang2004ncr,cali2000dti,lopez-toledo2006aoi,
barcelo2008lba,bellalta2009vtc,he2009srb,barcelo2010fcc,fang2011dlm,hui2011epp,barcelo2011tcf} \Az{(missing any?)} have proposed modifications to CSMA/CA (Carrier Sense Multiple Access with Collision Avoidance), the contention protocol used in the standard for sharing the medium. \Az{Despite the throughput improvement offered by all these works, none of the proposed modifications have yet been adapted by the standard.}

It is essential for a candidate CSMA/CA replacement to
\begin{itemize}
  \item provide performance advantages in terms of throughput and/or short term fairness;
  \item be backward compatible with the current implementation;
  \item support a large number of simultaneous contenders; and 
  \item be a simple evolution, in terms of implementation, to ease the transition and reduce the time to market (optional but desirable).
\end{itemize}

The aforementioned works can be categorized in three groups regarding the approach they use. \Az{In this paper, we will focus on the last of these three groups, proposing an additional enhancement to make it more adaptable to network size variations, and make the case for its adaptation by the standard.}

In the first group, represented by \cite{bharghavan1994map,wang2004ncr}, a throughput increase is obtained in saturation conditions by preventing the contention window from resetting to its minimum value after each successful transmission. However this performance gain comes at the expense of reduced short term fairness, since nodes that have recently gone through a sequence of consecutive collisions are forced to stay at a higher back-off stage and thus are further penalized by less frequent transmissions.

In the second group of works, exemplified here by \cite{cali2000dti,lopez-toledo2006aoi}, an accurate estimate of the number of contenders is used to adjust the contention parameters. This approach offers some throughput and fairness gains, but at the expense of increased implementation complexity. In addition, as the number of contenders is estimated relying on the number of collisions, the presence of channel errors further complicates the estimation \Az{[Complicates, or renders it inaccurate? Is there a work-around offered?]}.
Furthermore, there is a fundamental trade-off between the stability and the speed of reaction of the estimation.

When it comes to backward compatibility, neither of the aforementioned solutions are able to fairly share the medium with legacy devices.
In fact, these proposals are, generally speaking, less aggressive than the currently implemented CSMA/CA protocol. Consequently, in a hypothetical mixed network in which the new and old protocols coexist, the new stations would receive a smaller share of the available bandwidth.

A more important limitation of the two approaches exposed so far is that the throughput is bounded by that of CSMA/CA with optimal configuration \cite{cali2000dti,bianchi2000pai}.

The third group of solutions, which is the focus of the current paper, can deliver a throughput above the maximum attainable by CSMA/CA. This performance boost is achieved mainly by the use of deterministic backoff after successful transmissions, which reduces the chances of collisions for nodes that were successful in their previous transmission. Furthermore, under certain conditions, a collision-free operation can be reached. \Az{We will refer hereafter to the class of algorithms that use deterministic backoffs after successful transmissions as CSMA with Enhanced Collision Avoidance (CSMA/ECA).} This approach was first introduced in \cite{barcelo2008lba} \Az{(Does this contain analysis?)}, and later a more detailed analysis for both saturated and non-saturated conditions was presented in \cite{bellalta2009vtc}.
A more in-depth study is carried out in \cite{he2009srb}, including realistic elements such as the possibility of Clear Channel Assessment errors.
Different aspects of fairness are addressed in \cite{he2009srb,barcelo2010fcc,fang2011dlm}.
The performance in realistic channels, taking into account the Auto Rate Fallback mechanism, is evaluated in \cite{martorell2012pec,martorell2012tfl}.

@@

In the current paper, we present a completely distributed solution, called {\it backoff stage hysteresis} or simply hysteresis, to accommodate a large number of contenders in a fair collision-free fashion. We furthermore detail pseudo-code algorithms for three different types of CSMA/ECA and highlight how they can be easily implemented using simple modifications to the currently-used CSMA/CA algorithm. We will then present some performance evaluation, quantifying the performance gain that these three types of CSMA/ECA can achieve, both in terms of throughput and fairness.

\Az{-comment about backward compatibility and other items in the candidate requirements.}

@@

\begin{algorithm}
\While{the device is on}
{
  %\tcc{Initialize retransmission attempts.}
  $r \leftarrow 0$ ; $s \leftarrow 0$ \;
  %\tcc{Initialize backoff counter.}
  $b \leftarrow \mathcal{U}[0,2^s\rm{CW}_{min}-1]$\;
  \While{there is a packet to transmit}{
    %\tcc{Initialize $a$.}
    \Repeat{($r = R$) or (success)}{
      %\tcc{First, backoff.}
      \While{$b>0$}{
        wait 1 slot \;
        $b \leftarrow b-1$ \;
      }
      Attempt transmission of 1 packet \;
      \If{collision}{
        %\tcc{Random backoff.}
        $r \leftarrow r+1$ \;
        $s \leftarrow \min (s+1,S)$ \;
        $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min} -1]$\;
      }
    }
    \eIf{success}{
      %\tcc{Random backoff.}
      $r \leftarrow 0$ \;
      $s \leftarrow 0$ \;
      $b \leftarrow \mathcal{U}[0,2^{s}\rm{CW}_{min}-1]$\;
    }
    {
      Discard packet\;
      $r \leftarrow 0$ \;
      $s \leftarrow 0$ \;
      $b \leftarrow \mathcal{U}[0,2^s \rm{CW}_{min}-1]$\;
    }
  }
  Wait until there is a packet to transmit \;
}
\caption{CSMA/CA}
\label{alg:CSMA_CA}
\end{algorithm}

\begin{algorithm}
\While{the device is on}
{
  $r \leftarrow 0$ ; $s \leftarrow 0$ \;
  $b \leftarrow \mathcal{U}[0,2^s\rm{CW}_{min}-1]$\;
  \While{there is a packet to transmit}{
    %\tcc{Initialize $a$.}
    \Repeat{($r = R$) or (success)}{
      %\tcc{First, backoff.}
      \While{$b>0$}{
        wait 1 slot \;
        $b \leftarrow b-1$ \;
      }
      Attempt transmission of 1 packet \;
      \If{collision}
      {
        %\tcc{Random backoff.}
        $r \leftarrow r+1$ \;
        $s \leftarrow \min (s+1,S)$ \;
        $b \leftarrow \mathcal{U}[0, 2^s  \rm{CW}_{min} -1]$\;
      }
    }
    \eIf{success}{
      %\tcc{Random backoff.}
      $r \leftarrow 0$ \;
      $s \leftarrow 0$ \;
      $b \leftarrow (2^{s}\rm{CW}_{min})/2$\;
    }
    {
      Discard packet\;
      $r \leftarrow 0$ \;
      $s \leftarrow 0$ \;
      $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min}-1]$\;
    }
  }
  Wait until there is a packet to transmit \;
}
\caption{CSMA/ECA}
\label{alg:CSMA_ECA}
\end{algorithm}

\begin{algorithm}
\While{the device is on}
{
  $r \leftarrow 0$ ; $s \leftarrow 0$ \;
  $b \leftarrow \mathcal{U}[0,2^s\rm{CW}_{min}-1]$\;
  \While{there is a packet to transmit}{
    %\tcc{Initialize $a$.}
    \Repeat{($r = R$) or (success)}{
      %\tcc{First, backoff.}
      \While{$b>0$}{
        wait 1 slot \;
        $b \leftarrow b-1$ \;
      }
      Attempt transmission of 1 packet \;
      \If{collision}{
        %\tcc{Random backoff.}
        $r \leftarrow r+1$ \;
        $s \leftarrow \min (s+1,S)$ \;
        $b \leftarrow \mathcal{U}[0, 2^s  \rm{CW}_{min} -1]$\;
      }
    }
    \eIf{success}{
      %\tcc{Random backoff.}
      $r \leftarrow 0$ \;
      %$s \leftarrow 0$ \;
      $b \leftarrow (2^{s}\rm{CW}_{min})/2$\;
    }
    {
      Discard packet\;
      $r \leftarrow 0$ \;
      %$s \leftarrow 0$ \;
      $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min}-1]$\;
    }
  }
  Wait until there is a packet to transmit \;
}
\vspace{0.2cm}
\caption{CSMA/ECA with hysteresis}
\label{alg:CSMA_ECA_hist}
\end{algorithm}
